Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-02-02 paper.tex
some more text

>>>>

The core concepts involved in computational drug discovery can be
reduced to genes, proteins, pathways and small molecules interacting
with them. Computational representations of these concepts and the
methods to be manipulate them have been implemented in a variety of
software tools. A comprehensive computational drug discovery platform
will include a number of these tools, as it is rare that a single tool
addresses all aspects.

In this sense, the R environment presents a robust platform for
computational drug discovery, especially the stages up to and
including lead optimization. The Bioconductor set of packages provide
a plethora of tools to analyze a variety of ``omics'' data. Examples
include packages that support gene expression (microarray or RNAseq)
analyses, tools for differential expression analysis, gene and pathway
enrichment analysis and so on.

In contrast, support for small molecule data is limited. Currently,
there are two packages that focus on representing and manipulating
chemical structures - \texttt{ChemmineR} \cite{Cao:2008fj} and \rcdk
\cite{Guha:2007aa}. Both expose cheminformatics functionality
implemented in an underlying Java library (JOELib for ChemmineR and
the CDK for \rcdk). In this section we focus on the use of \rcdk as a
means to manipulate and analyze chemical structure data.

The \rcdk package supports a variety of chemical structure file
formats and presents the CDK Java API as a set of idiomatic R
functions. That is, rather than manually specify fully qualified Java
packages or resort to writing ``Java in R'' the \rcdk package wraps
much of the CDK API as ordinary R functions. Thus the atoms in a
molecule are represented as a set of atom objects. Importantly, atoms
(and other objects such as bonds and molecules) are stored as
references to Java objects. While they can be manipulated directly, it
is unwieldy it is recommended that one uses methods provided by the
\rcdk package.
